---
title: "STAA57 W21 - Project Proposal"
author:  "Group 11 (Alice Huang, Dominic Ma, Jalal Kassab, Vanshika Virmani)"
output:
  pdf_document:
    df_print: paged
urlcolor: blue
geometry: margin=1 cm
---
Link to the shared RStudio Cloud project that created this report: https://rstudio.cloud/spaces/115177/project/2132391    
The final R code can be found in the Proposal_Template_FINAL.Rmd file.

### Objectives of Our Investigations

We are interested in efficient pricing of flight training courses. So, we plan to analyse information useful to the Durham Flight Training Centre, in terms of the costs that they must incur for offering the training, and how they should conduct training with the exercises and aircrafts that will maximize the chances for success in training aviation students. 

To do so, we are defining student ‘success’ by determining whether he/she can fly solo at least once during the course of their training. 

For the cost analysis for the Durham Flight Centre, we would like to estimate: 

1. How much money the training centre has spent on fuel and maintenance for different aircraft models over the past few years. 
+ From Durham Flight Centre’s website, we know on average how much the centre was charging for flying each plane per hour. This will help us determine the profit margins for the flight centre. 
2. Whether the number of flight hours have differed across different times of the year. And whether some planes are more in demand during certain months. 
+ This will help us determine how the demand varies for the flights throughout the year. If we’re able to see that certain planes are more in demand during certain months, then we can analyse how the flight centre can adjust the prices for different types of trainings to increase profit.
3. How much money it costs to train the average student until they can safely take off, land and fly solo. 

For answering questions in regards to the training schedule for students, we would like to build a “profile” for students. In order to determine how the centre can maximize student success, we want to know: 

4. What are the types of exercises that students who were able to fly solo at least once completed that set them apart from students who weren’t able to fly solo at all? 
5. Which types of planes did the ‘successful’ students use to complete certain exercises? This will help us map out information for the Durham Training Centre, as to which planes they should allocate for which exercises to maximise student success in training. 

By further analysing the types of exercises and the aircraft models suited for those exercises, we’d be able to determine how the training centre should schedule certain exercises for students. More student success can in return help the centre attract more customers. 

### Data Analysis Plan

1. To estimate how much money the flight training centre has spent on fuel and maintenance costs, we will acquire a range of estimates for the hourly fuel consumption and maintenance fees of each aircraft and we will add up the duration of time spent on training with each type of plane. This will be done by summing up the duration column for each type of plane. This will provide us with a graph that has the x axis being the types of planes, and the y axis being the total cost of flying that plane. 
To estimate revenue from the trainings, we will find values of how much they charge each student for flying the different types of planes from the training centre’s website, and multiply the values with the total duration of flights for each plane. We will then subtract the cost values for each plane for the revenue values to estimate the flight centre’s profit margin on each type of plane. 

2. We will group the duration of flights by month for each year, and sum up the number of hours. This gives us the plots with how long the students trained for in each month of the corresponding years. 

3. To estimate the cost for the “average student” in the training centre, we will group the data by different student IDs and plane types and then find the average number of hours spent in flight training by each student for each plane type. We will then multiply these values with the average costs from above, and create a graph with the x axis as the types of planes and y values as the average cost for training a student. 

4. We will find the students who flew solo at least once by finding which students’ training types contained the substring “solo”. And we will filter out the students who didn’t fit this criteria. We will then map the frequency of exercises completed by each student, and add the frequencies together for all exercises. This graph will show us the most frequent exercises that "successful" students completed. 

5. Repeat the same process with filtering out students who got to fly “solo”. We will then add up the duration of hours for these students with the types of planes used to see if they used one plane more frequently than the other. We will also group the most frequency exercises completed from above by the type of airplane to see which planes are better suited for which exercise. 

### Data Sources

We intend to use a dataset provided by our client, the Durham Flight Training Centre. This dataset logs the exercises students completed with different instructors on different dates. It also specifies the durations of the training sessions, whether the exercises were completed solo or with an instructor, and which aircrafts were used. The dataset only spans the years 2002, 2015-2020, so we can only observe patterns and compare trends over the span of five years, and it may not be the most suitable for generalizing trends to longer periods of time. Since there is a gap of 13 years between 2002 and 2015, we didn’t think it’d make sense to do a lot of analysis on the data from 2002 as a lot of factors such as aircraft condition, market conditions, etc. could have changed in the meantime.

The data we gathered on the aircrafts’ ownership costs (C-152, C-172, and C-150) was obtained from aopa.org, the website of the "Aircraft Owners and Pilots Association", and https://cessna150152club.org/, the website of the "Cessna 150-152 Club", a membership club and nonprofit dedicated to educating prospective and current pilots about the Cessna-152 and Cessna-150. We believe these organizations’ numbers should be reliable since these organizations and communities have access to active users of Cessna-172, Cessna-152, Cessna-150 planes. 

However, the data gathered from the "Cessna 150-152 Club" may be a little biased since the entire organization is based on those two planes and the content on the website seems to feature more positive anecdotal experiences with the two planes. Furthermore, the ownership club may not include as much information for users who did not pay their club membership fee, and others who are not active on their internet forums. Due to these concerns, we checked if their numbers for ownership costs were consistent with other websites, and they were, so we chose this as a data source. 

The website for the "Aircraft Owners and Pilots Association" helped us find out the fuel efficiency and price for all three planes. But we had to decide whether we should use the average used price of the planes or the price of the "reimagined" planes, which have been overhauled, repainted and serviced. We ended up using the price of the "reimagined" planes since it gave us a better idea about the actual price difference without any external variables.

Both the "Aircraft Owners and Pilots Association" and "Cessna 150-152 Club" are based in the United States so costs, regulations and experiences may be biased towards English-speaking American users of Cessna-172, Cessna-152 and Cessna-150 planes. Our client is based in Southern Canada, close to the US border, but we believe certain costs like a plane's fuel burned per hour, should not depend on region. Insurance and inspection costs may vary by region, however.

### Importing and Formatting the Data in R

```{r, include=FALSE, warning=FALSE}
source(file = "preprocess_data.R")
library(tidyverse)
```

We found estimates for fuel and operating costs from the “Aircraft Owners and Pilots Association'' and the “Cessna 150-152 Club”. These seemed to be the most accurate numbers we could find. These websites did not post their numbers as datasets with csv, txt, xlsx, xml, or other formats that are convenient to work with. In particular, on the “Cessna 150-152 Club” website's "Members Only" pages, it says that servers containing their data were hacked, so the organization had to pull their data offline and is now in the process of "building a modern and secure database structure". Since we could not find a pre-made dataset, we made a csv file, and entered the hourly estimates of various costs for operating and maintaining the aircraft in the tidy data format. We put this csv file in our data folder with the rest of our project. Wherever the sites provided a range of values, we just took the median for our computations as the median is a good indicator of the "middle value" in a distribution and is less likely to be affected by outliers. 

```{r , include = FALSE, warning=FALSE}
planecosts = read_csv("/cloud/project/data/planescosts.csv")
planecosts
```

An issue that we ran into while trying to organize exercise data was the fact that the elements were stored as string rather than list objects. Judging from the last few lines of the preprocessed data script, this was not meant to be the case. Although we could not figure out why those lines didn’t work, we managed to work around it by creating a copied dataframe with the changes we wanted. In hindsight, it’s likely because dataframes are immutable. This will be useful for future analyses on the frequency of Exercises completed.

```{r, include = FALSE, warning=FALSE}
my_clean_data <- clean_data %>% 
  distinct( Session_ID, .keep_all = T) %>% 
  mutate(Exercises = str_split(Exercises, ","))
```

Furthermore, when we were plotting the months of the year vs the duration of training hours, we found a year corresponding to the value “201”, which we speculated was a typing error. So when graphing the data, we only graphed for years past “2010”. We also found an instance where a plane was entered as "C152" while its counterparts were entered as "C-152". We ensured all Cessna-152 planes were entered as "C-152" for consistency. If we run into more typos or strange data values, we will communicate with Durham Flight Training Centre, and change or filter values as appropriate.

````{r , include = FALSE, warning=FALSE}
clean_data$Aircraft[clean_data$Aircraft == "C152"] <- "C-152"
```

### Preliminary Results

We wish to estimate the operating and maintenance cost the training centre has spent on each type of aircraft. Thus we are interested in the total amount of time students spent with each of the different types of aircrafts. We previously found estimates for how much fuel each plane burns, in gallons per hour, the unit pilots conventionally use to measure hourly costs. Computing the hours spent on flying the different aircrafts will help us estimate how much fuel was consumed by each type of aircraft, and in turn, how much money was spent on fuel for each aircraft.
Due to space limitations, we will not show all the code and the numbers we calculated, but they can be found in our RStudio Project Folder.

```{r , include = FALSE, warning=FALSE}
#Here is a table that shows the total time the centre spent on each type of aircraft from 2015-2020.
clean_data %>% 
  filter(Year >= 2010) %>%
  select(Aircraft, Duration, Training_Type) %>%
  group_by(Aircraft) %>%
  summarize(total_time_aircraft = sum(Duration))
```

```{r , include = FALSE, warning=FALSE}
# Here is code that plots the aircraft against the duration of time spent on that aircraft, with colors to denote solo and dual flights.
clean_data %>% ggplot(aes(x = Aircraft, y = Duration, fill = Training_Type)) + geom_bar(stat="identity")
clean_data %>% ggplot(aes(x = Aircraft, y = Duration, fill = Training_Type)) + geom_bar(stat="identity") + facet_wrap(Year~., scales = "free") + theme(legend.position = "bottom")
```


```{r , include = FALSE, warning=FALSE}
# Here is a table that shows the total fuel consumed per aircraft over the years 2015-2020.
clean_data %>% left_join(planecosts, by="Aircraft") %>%
  mutate(Fuel_Cost_Gallons_Session = Duration*Fuel_Cost_Per_Hour) -> clean_data_withcosts

clean_data_withcosts %>% 
  filter(Year >= 2010) %>%
  group_by(Aircraft) %>%
  summarise(total_fuel_consumed = sum(Fuel_Cost_Gallons_Session)) -> total_aircraft_costs

total_aircraft_costs
```

```{r ,include = FALSE, warning=FALSE}
# Here is a table that shows the total fuel consumed per aircraft over each year in the range 2015-2020.
clean_data_withcosts %>% 
  filter(Year >= 2010) %>%
  group_by(Year, Aircraft) %>%
  summarise(total_fuel_consumed = sum(Fuel_Cost_Gallons_Session)) -> total_aircraft_costs_year

total_aircraft_costs_year
```

Here are some graphs that show the total fuel consumed per aircraft over each year in the range 2015-2020. 

````{r , echo = FALSE, warning=FALSE}
total_aircraft_costs_year %>% ggplot(aes(x = Aircraft, y = total_fuel_consumed)) + geom_bar(stat = "identity") + facet_wrap(.~Year, scales = "free") + theme(legend.position = "bottom")  + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + ggtitle("Total Fuel Consumed on Different Aircrafts For Years 2015-2020") + ylab("total fuel consumed (gallons per hour)")
```

We found that over the years 2015-2020, the training centre’s Cessna-172 aircrafts consumed around 15856.68 gallons of fuel in total. The Cessna-152 aircrafts consumed around 6021.92 gallons of fuel in total. The Cessna-150 aircrafts consumed around 997.92 gallons of fuel in total. This is consistent with the fact that over the span of those 5 years, the Cessna-172 was used for training more often, and its hourly fuel cost is more expensive than that of the Cessna-152 and Cessna-150. It is interesting to note that the training centre spent nearly 2 times the amount of time on Cessna-172 than the Cessna-152, but it spent close to 3 times the amount of fuel on Cessna-172 than the Cessna-152.

Now when we looked at the fuel consumption per year, we saw that in 2019-2020, the Cessna-172 consumed much more fuel than the Cessna-152. In 2019, the C-172 consumed approximately 6622.56 gallons of fuel while the C-152 consumed approximately 2009.95 gallons of fuel. In January-March, June-September 2020, the C-172 consumed approximately 8106.84 gallons of fuel while the C-152 consumed approximately 2940.20 gallons of fuel. Again, this is in line with the frequency of Cessna-172 training sessions during those years. During the years 2015-2018, the flight centre did not use the Cessna-172 as often so the C-172s did not consume as much fuel.

A limitation of our current analysis is that we assumed the gallons of fuel consumed per hour remained constant over 2015-2020 regardless of aircraft age and condition, nature of Exercise, market conditions for fuel pricing, and other factors. We’d like to take our current analysis a step further if possible. If we can find reliable datasets for how aircraft fuel costs varied over 2015-2020 in Ontario, we’d like to further improve our estimates of fuel costs. Then we’d be able to translate gallons consumed per hour to monetary costs, and further inform our client about pricing decisions.

We made some graphs graphing the months of the year against the total duration of training session for those months. We made separate graphs for each year from 2016-2020, and analyzed them. The code for the separate graphs can be seen in our R Studio Cloud Project R Code. Due to space limitations, we'll just show the graph with the different months of all years in range 2016-2020.

```{r, include=FALSE, warning=FALSE}
clean_data %>% select(Aircraft, Duration, Training_Type, Month, Year) %>%
  filter( Month <= 12) %>%
  filter(Year == 2016) %>%
  group_by(Aircraft, Month, Year) %>%
  summarize(total_time_per_month = sum(Duration)) %>%
  ggplot(aes(x = factor(Month), y = total_time_per_month, fill = factor(Month))) + geom_bar(stat="identity") + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + ggtitle("Months in 2016 vs Duration of Flight Hours") + scale_fill_discrete(name = "Months",
labels = c("January", "February", "March" ,"April", "May", "June", "July", "August", "September", "October", "November"))

```

```{r, include=FALSE, warning=FALSE}
clean_data %>% select(Aircraft, Duration, Training_Type, Month, Year) %>%
  filter( Month <= 12) %>%
  filter(Year == 2017) %>%
  group_by(Aircraft, Month, Year) %>%
  summarize(total_time_per_month = sum(Duration)) %>%
  ggplot(aes(x = factor(Month), y = total_time_per_month, fill = factor(Month))) + geom_bar(stat="identity") + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + ggtitle("Months in 2017 vs Duration of Flight Hours") + scale_fill_discrete(name = "Months",
labels = c("January", "February", "March" , "May", "June", "July", "August", "September", "October", "November", "December"))

```

```{r, include=FALSE, warning=FALSE}
clean_data %>% select(Aircraft, Duration, Training_Type, Month, Year) %>%
  filter( Month <= 12) %>%
  filter(Year == 2018) %>%
  group_by(Aircraft, Month, Year) %>%
  summarize(total_time_per_month = sum(Duration)) %>%
  ggplot(aes(x = factor(Month), y = total_time_per_month, fill = factor(Month))) + geom_bar(stat="identity") + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + ggtitle("Months in 2018 vs Duration of Flight Hours") + scale_fill_discrete(name = "Months",
labels = c("February", "March" ,"April", "May", "June", "July", "August", "September", "October", "November", "December"))

```

```{r, include=FALSE, warning=FALSE}
clean_data %>% select(Aircraft, Duration, Training_Type, Month, Year) %>%
  filter( Month <= 12) %>%
  filter(Year == 2019) %>%
  group_by(Aircraft, Month, Year) %>%
  summarize(total_time_per_month = sum(Duration)) %>%
  ggplot(aes(x = factor(Month), y = total_time_per_month, fill = factor(Month))) + geom_bar(stat="identity") + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + ggtitle("Months in 2019 vs Duration of Flight Hours") + scale_fill_discrete(name = "Months",
labels = c("January", "February", "March" ,"April", "May", "June", "July", "August", "September", "October", "November", "December"))

```

```{r, include=FALSE, warning=FALSE}
clean_data %>% select(Aircraft, Duration, Training_Type, Month, Year) %>%
  filter( Month <= 12) %>%
  filter(Year == 2020) %>%
  group_by(Aircraft, Month, Year) %>%
  summarize(total_time_per_month = sum(Duration)) %>%
  ggplot(aes(x = factor(Month), y = total_time_per_month, fill = factor(Month))) + geom_bar(stat="identity") + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + ggtitle("Months in 2020 vs Duration of Flight Hours") + scale_fill_discrete(name = "Months",
labels = c("January", "February", "March" , "June", "July", "August", "September", "December"))

```

```{r, echo = FALSE, warning = FALSE}
clean_data %>% select(Aircraft, Duration, Training_Type, Month, Year) %>%
  filter(Aircraft == "C-152") %>%
  filter(Year> 2010) %>%
  mutate( Month_Year = str_c(Year, Month, sep = " - ")) %>%
  group_by(Month_Year) %>%
  summarize(total_time_per_month = sum(Duration)) %>%
  ggplot(aes(x = factor(Month_Year), y = total_time_per_month, fill = Month_Year)) + geom_bar(stat="identity") + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + theme(legend.position = "none") + ggtitle("Year_Month vs Duration of Flight Hours per Month") + ylab("Total Flying Time Per Month (hours)") + xlab("Year - Month")
  
```

Note that 2020 only has data for 7 months because of the lockdown from the pandemic. Also, 2018 has no data for January. That is why we thought it would be unfair to find the aggregate of duration for each month across all years (eg. add up duration for 2016 January, 2017 January, etc), because the data would be slightly skewed. So, we thought about representing the information as 5 separate graphs from each year. 

We notice that the training duration generally increases during the months of July, August, September, October and November, usually peaking around August. This may be due to the fact that those months tend to be warmer, and  invite clearer skies which are an important consideration when flying. We think it'd be interesting to compare the training duration with more detailed weather data on temperatures and precipitation.

Moreover, we see that from 2016 to 2017 the hours rarely surpass 10, and increase to peak around 60 in 2018. In 2019 it reaches up to 180, but oddly enough 2020 seems to have the highest peak of above 400 flight hours. So even during a pandemic, the flight center was able to maintain growth and bring in more customers. This steady growth in flight hours, which is synonymous to the demand for training, has been increasing over the years, as shown in our graphs. But to help tackle the slump in demand that the flight center sees every winter/spring, they could decrease prices for the training courses (of course while staying above the fixed + variable costs), which in turn could increase demand and possibly increase growth over the years. 

We made two bar graphs which can be seen on the next page. The first shows the students who were able to fly solo at least once, their aircraft choice, and the total time flown on those aircrafts. The second shows the same information for students who were NOT able to fly solo at least once. These graphs show us the “successful” student’s behaviour and their aircraft of choice.

When we compared the graphs for students who have flown solo and those who did not, we saw that those who have flown solo prefer sticking to one aircraft and rarely switch between the two. We believe that that students who stick to their preferred aircraft end up becoming more comfortable with it and mastering it at a faster pace. Meanwhile students who tend to switch between 2 or 3 aircrafts might struggle with adjusting to different aircraft, and end up taking a longer time to perfect their craft.

Our goal is to try to find a benchmark of exercises that all “successful” students do that might show types of exercises that lead to greater student success. We are still working on a graph that shows the exercises on the x axis and their frequency on the y axis for successful students. 

```{r echo=FALSE, warning=FALSE}
clean_data %>%
  select(Student_ID, Aircraft, Duration, Instructor_ID, Training_Type) %>%
  mutate(solo = str_detect(Training_Type, "solo")) %>%
  group_by(Student_ID, Aircraft) %>%
  summarize( successful = sum(solo), total_time = sum(Duration)) %>%
  filter(successful > 0) %>%
  ggplot(aes(x = factor(Student_ID), y = total_time, fill = Aircraft)) + geom_bar(stat = "identity") + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + ggtitle("Students Who Have Flown Solo vs. Flight Time on C-152, C-172") + ylab("Total Flying Time Per Month (hours)") + xlab("Student ID")

```

```{r echo=FALSE, warning=FALSE}
clean_data %>%
  select(Student_ID, Aircraft, Duration, Instructor_ID, Training_Type) %>%
  mutate(solo = str_detect(Training_Type, "solo")) %>%
  group_by(Student_ID, Aircraft) %>%
  summarize( successful = sum(solo), total_time = sum(Duration)) %>%
  filter(successful == 0) %>%
  ggplot(aes(x = factor(Student_ID), y = total_time, fill = Aircraft)) + geom_bar(stat = "identity") + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + ggtitle("Students Who Have Not Flown Solo vs. Flight Time on Various Aircrafts") + ylab("Total Flying Time Per Month (hours)") + xlab("Student ID")

```