---
title: "STAA57 W21 - Project Proposal"
author:  "Group 11 (Alice Huang, Dominic Ma, Jalal Kassab, Vanshika Virmani)"
output:
  pdf_document:
    df_print: paged
urlcolor: blue
---
***

This is a draft file which I used for testing things out. Final report is in the file
labelled Proposal_Template_FINAL.Rmd

Link to the share RStudio Cloud project that created this report: e.g.  
https://rstudio.cloud/spaces/115177/project/2132391 (where XYZ is the project #)       

#### Analysis Plan 

1. Specify the questions you will address. Describe the general topic of your investigation, and state specific questions you will address. Include any relevant information.

2. Specify your data analysis plan. Describe (in words) how you will address these questions using data.

3. Specify your data sources, and the type of information you will use (for external data, provide links/references). For each data source, describe the variables and observations used in your analysis. Identify any potential issues (e.g. bias). 

4. Provide the R code that imports the data into R, and formats them appropriately (this can go to the appendix if it is too lengthy). Submit a copy of your external data files, if any. 
#### Data 
```{r, include=FALSE, warning=FALSE}
source(file = "preprocess_data.R")
library(tidyverse)
```

#### Preliminary Results

5. Create at least three data summaries/visualisations which are relevant to your questions, and comment on your results. 

```{r}
clean_data %>%
  select(Student_ID, Aircraft, Duration) %>%
  group_by(Student_ID, Aircraft) %>%
  summarize( total_time = sum(Duration)) -> aircraft_time_data

aircraft_time_data

aircraft_time_data %>%
  ggplot(aes(x = Student_ID, y = total_time, fill = Aircraft)) + geom_bar(stat = "identity", position = "stack")
```


```{r}
clean_data %>%
  select(Student_ID, Duration, Training_Type) %>%
  group_by(Student_ID, Training_Type) %>%
  summarize(total_training_time = sum(Duration)) -> solo_dual_data

solo_dual_data

solo_dual_data %>% ggplot(aes(x=Student_ID, y=total_training_time, fill=Training_Type)) + geom_bar(stat = "identity")
```

```{r}
clean_data %>% select(Aircraft, Duration, Training_Type) %>%
  group_by(Aircraft) %>%
  summarize(total_time_aircraft = sum(Duration))

clean_data %>% ggplot(aes(x = Aircraft, y = Duration, fill = Training_Type)) + geom_bar(stat="identity")
```
```{r}

planecosts = read_csv("/cloud/project/data/planescosts.csv")
planecosts
```



```{r}
clean_data %>% select(Student_ID) %>%
  summarize(Student_ID, Duration, total_hours = sum(Duration)) -> students2

students2

clean_data %>% filter(Training_Type == "CC_solo"| Training_Type == "LF_solo") %>%
  distinct(Student_ID) -> solo_students

solo_students

solo_students %>% ggplot(aes(x= Student_ID))
```

```{r}
# count how many times student 4 finished a different exercise

clean_data %>% select(Student_ID, Exercises) %>%
  filter(Student_ID == 4) -> student_4_ex

student_4_ex

student_4_ex %>% count("23")
```

```{r}
#This is a function that computes the fuel cost for an aircraft given its model (one of C-172, C-152, C-150)
# and the number of hours it flew (hours)
planecosts %>% filter(Aircraft == "C-172") %>% select(Fuel_Cost_Per_Hour) -> hourly_cost_c172
planecosts %>% filter(Aircraft == "C-152") %>% select(Fuel_Cost_Per_Hour) -> hourly_cost_c152
planecosts %>% filter(Aircraft == "C-150") %>% select(Fuel_Cost_Per_Hour) -> hourly_cost_c150

fuel_cost = function(plane, hours){
  result = NA
  if(plane == "C-172"){
    result = hours*hourly_cost_c172[[1]]
  } 
  else if(plane == "C-152"){
    result = hours*hourly_cost_c152[[1]]
  } 
  else if(plane == "C-150"){
    result = hours*hourly_cost_c150[[1]]
  }
  return(result)
}
```

```{r}
clean_data %>% left_join(planecosts, by="Aircraft") %>%
  mutate(Fuel_Cost_Gallons_Session = Duration*Fuel_Cost_Per_Hour) -> clean_data_withcosts

clean_data_withcosts %>% group_by(Aircraft) %>%
  summarise(total_costs = sum(Fuel_Cost_Gallons_Session)) %>%
  ggplot(aes(x = Aircraft, y = total_costs)) + geom_bar(stat = "identity")
```
```{r , echo = FALSE, warning=FALSE}
clean_data %>%
  select(Student_ID, Duration, Training_Type) %>%
  group_by(Student_ID, Training_Type) %>%
  summarize(total_training_time = sum(Duration)) -> solo_dual_data

solo_dual_data

solo_dual_data %>% ggplot(aes(x=Student_ID, y=total_training_time, fill=Training_Type)) + geom_bar(stat = "identity")
```


```{r , echo = FALSE, warning=FALSE}
clean_data %>% select(Aircraft, Duration, Training_Type) %>%
  group_by(Aircraft) %>%
  summarize(total_time_aircraft = sum(Duration))

clean_data %>% ggplot(aes(x = Aircraft, y = Duration, fill = Training_Type)) + geom_bar(stat="identity")
clean_data %>% ggplot(aes(x = Aircraft, y = Duration, fill = Training_Type)) + geom_bar(stat="identity") + facet_wrap(Year~., scales = "free") + theme(legend.position = "bottom")
```

*************************
```{r}
my_clean_data %>% filter(Student_ID %in% c(1,6,54,121,41,8,79,87,93,44,4,91,27)) %>%
  mutate(New_Duration = ifelse(is.na(Duration), 0, Duration)) %>%
  group_by(Student_ID) %>%
  summarize(sum(New_Duration))
```

WARNING: does not account for issue with GROUND school students!!
```{r}
clean_data%>%
  filter(Aircraft == "C-152"|| "C-172")%>%
  mutate(Revenue = ifelse(Aircraft =="C-152", 135*ceiling(Duration), 155*ceiling(Duration)))%>%
  mutate(Instructor_costs = ifelse(Training_Type == "LF_dual"|Aircraft == "GROUND", Duration*65, 0)) %>% 
  filter(Year >= 2018) %>%
  mutate( Month_Year = str_c(Year, Month, sep = " - ")) %>%
  group_by(Month_Year) %>%
  summarize(total_revenue_per_month = sum(Revenue)) %>% 
  ggplot(aes(x = factor(Month_Year), y = total_revenue_per_month, fill = Month_Year)) + geom_bar(stat="identity") + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + theme(legend.position = "none") + ggtitle("Year_Month vs Monthly Revenue") + ylab("Monthly Revenue ($CAD)") + xlab("Year - Month")
```

**********************************************
*********************************************

## Monthly Revenue Code

```{r}
hourly_c150c152_rental_rate = 135
hourly_c172_rental_rate = 155

my_cd%>%
  filter(Aircraft %in% c("C-152", "C-172", "C-150"))%>%
  mutate(Revenue_Session = ifelse(Aircraft =="C-172", hourly_c172_rental_rate*ceiling(Duration), hourly_c150c152_rental_rate*ceiling(Duration)))%>%
  mutate(Instructor_costs = ifelse(Training_Type == "LF_dual"|Aircraft == "GROUND", Duration*65, 0)) %>%
  filter(Year >= 2018) %>%
  mutate( Month_Year = str_c(Year, Month, sep = " - ")) %>%
  group_by(Month_Year) %>%
  summarize(total_revenue_per_month = sum(Revenue_Session)) -> lower_bound_month_revenue

(lower_bound_month_revenue %>% mutate(Lower_or_Upper = "Lower") -> lower_bound_month_revenue)

# ggplot(aes(x = factor(Month_Year), y = total_revenue_per_month, fill = Month_Year)) + geom_bar(stat="identity") + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + theme(legend.position = "none") + ggtitle("Year_Month vs Monthly Revenue") + ylab("Monthly Revenue ($CAD)") + xlab("Year - Month")
```
```{r}
# cost for ground session

package_5hr_revenue = 1575
c172_package_upgrade = 100

(ground_revenue5hr = package_5hr_revenue - 5*hourly_c150c152_rental_rate)

package_10hr_revenue = 2575

(ground_revenue10hr = package_10hr_revenue - 10*hourly_c150c152_rental_rate)
```


```{r}
my_cd%>%
  filter(Aircraft %in% c("C-152", "C-172", "C-150", "GROUND"))%>%
  mutate(Revenue_Session2 = ifelse(Aircraft =="C-172", hourly_c172_rental_rate*ceiling(Duration), ifelse(Aircraft=="GROUND", ground_revenue10hr, hourly_c150c152_rental_rate*ceiling(Duration))))%>%
  filter(Year >= 2018) %>%
  mutate( Month_Year = str_c(Year, Month, sep = " - ")) %>%
  group_by(Month_Year) %>%
  summarize(total_revenue_per_month = sum(Revenue_Session2)) -> upper_bound_month_revenue

(upper_bound_month_revenue %>% mutate(Lower_or_Upper = "Upper") -> upper_bound_month_revenue)
```

```{r}
bind_rows(lower_bound_month_revenue, upper_bound_month_revenue) -> monthly_revenue_both
monthly_revenue_both
```

```{r message=FALSE, warning=FALSE}
ggplot(monthly_revenue_both, aes(x=Month_Year, y = total_revenue_per_month)) + geom_bar(stat = "identity") + facet_grid(.~ Lower_or_Upper) + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + theme(legend.position = "none") + ggtitle("Year_Month vs Monthly Revenue") + ylab("Monthly Revenue ($CAD)") + xlab("Year - Month")
```

## Monthly Fuel Costs

````{r , echo = FALSE, warning=FALSE}
my_cd %>% mutate(Gallons_Fuel_Session = ifelse(Aircraft == "C-172", 8.4*Duration, 
                                                             ifelse(Aircraft == "C-152", 6.1*Duration, 5.6*Duration))) %>%
  mutate(Price_Fuel_Session = Gallons_Fuel_Session*(0.01/0.2199692)) %>%
  filter(Year >= 2018) %>%
  mutate( Month_Year = str_c(Year, Month, sep = " - ")) %>%
  group_by(Month_Year) %>%
  summarize(total_monthly_fuel = sum(Price_Fuel_Session)) %>%
  ggplot(aes(x = Month_Year, y = total_monthly_fuel)) + geom_bar(stat = "identity") + theme(legend.position = "bottom")  + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + ggtitle("Monthly Costs of Fuel for Years 2018-2020") + ylab("Monthly Fuel Cost ($CAD)")
```